---
title: "LAND PRODUCTIVITY DYNAMICS"
author: ""
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: 
  word_document:
    reference_docx: LPD_MS_styles.docx
    toc: true #table of content true
    toc_depth: 3  #up to three depths of headings (specified by #, ## and ###)
    #number_sections: true  #number sections at each table header
    #theme: united  #options for theme
    #highlight: tango  #syntax highlighting style
    #css: my.css   #custom css, should be in same folder
bibliography: lpd_biblio.bib
#csl: methods-in-ecology-and-evolution.csl
---


```{r setup, include = FALSE, results='asis'}
library(knitr)
library(pander)
library(captioner)
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
if(Sys.info()[4] == "D01RI1700371"){
  source("E:\\rotllxa\\LPD\\LPD/00_settings.R")
}else if(Sys.info()[4] == "h05-wad.ies.jrc.it"){
  source("/home/rotllxa/LPD/LPD/00_settings.R")
}else if(Sys.info()[4] == "MacBook-MacBook-Pro-de-Xavier.local"){
  source("/Users/xavi_rp/Documents/D6_LPD/LPD/00_settings.R")
}

table_num <- captioner::captioner(prefix = "Table")
fig_num <- captioner::captioner(prefix = "Figure")
#ts <- 0
```



# Abstract

[@ivits_assessing_2016]


# Introduction (Background)

```{r echo=FALSE}

tb1 <- table_num(name = "t1", caption = "bla bla bla")
 
```
 
`r table_num("t1", display = "cite")`

a) general introduction
UNCCD, SDG 15.3, Indicator


b) Land Productivity Dynamics Map: 

Include here a diagram of the whole process
Phenological variables (Phenolo or others), Net Primary Productivity, etc (include here Eva's graph: Figure 3: Schematic representation of the main phenological and productivity variables calculated by Phenolo (reprint from Ivits et al., 2013a).)

Land Productivity Dynamics Map = Long Term Change Map + Current Status Map of Land-Productivity

Qualitative indicator...






# Phenolo
Should this part be removed?? I think so...



# Long-Term Change Map of Land-Productivity



## Steadiness Index for Standing Biomass

```{r echo=FALSE}
fig1 <- fig_num(name = "f1", caption = "ble ble ble")
slp_rstr <- paste0(path2saveTests, "/slope_rstr.jpg")
```


`r fig_num("f1", display = "cite")`

![`r fig1`](`r slp_rstr`)






## Combining Steadiness Index with Baseline Levels of Standing Biomass


## Standing Biomass State Change


## Land-Productivity Long Term Change Map

```{r echo=FALSE}
fig5 <- fig_num(name = "f5", caption = "blu blu blu")
LandProd_change <- paste0(path2saveTests, "/LandProd_change.jpg")
```


![`r fig5`](`r LandProd_change`)


# Current Status Map of Land-Productivity


# Long Term Change Map + Current Status Map of Land-Productivity





# References

